---
title: "Accounts"
output:
  html_document: default
  word_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r echo=FALSE, message=FALSE}
library(lubridate)
library(data.table)
library(dplyr)
library(formattable)
library(tidyr)
```

## Bank

The Bank Account records transactions in the following categories:

- Sales
- Purchases
- Payments to Revenue
- Repayments of Director's Expenses
- Bank Charges

```{r echo=FALSE}
year <- 2025
docs <- paste("~/Documents/company/", year, "/data/", sep="")
date.fmt <- "%d %b %Y"
bank_doc <- paste(docs, "bank.csv", sep="")

bank <- read.csv(file = bank_doc, header=TRUE, sep=",")
bank$date <- format(as.Date(bank$date), date.fmt)
bank$amount <- accounting(bank$amount)
bank$balance <- currency(cumsum(bank$amount), symbol = "€")

bank <- bank[, c("date", "comment", "amount", "balance")]

formattable(bank, align=c("l", "l", "r", "r"))
```
## Purchases

Purchases are payments for goods and services made directly from the Bank Account.

```{r echo=FALSE}
purchases_doc <- paste(docs, "purchases.csv", sep="")
purchases <- read.csv(file = purchases_doc, header=TRUE, sep=",")
purchases$date <- format(as.Date(purchases$date), date.fmt)
purchases$allowed[is.na(purchases$allowed)] <- 100
purchases$allowed <- purchases$allowed / 100
purchases$expense <- currency(purchases$amount * purchases$allowed, symbol = "€")
purchases$amount = currency(purchases$amount, symbol = "€")
purchases$balance = currency(cumsum(purchases$expense), symbol = "€")
purchases$type = "P"

p <- purchases[, c("date", "description", "amount", "allowed", "expense", "balance")]

formattable(p, list(allowed = percent), align=c("l", "l", "r", "r", "r", "r"))
```
## Director's Expenses

Director's Expenses are payments for goods and service made from a Director's personal account.

The company may reimburse part or all of the payment, using the "allowed" column. This is a percentage,
which defaults to 100.

It is convenient to reimburse the Director when VAT is paid to Revenue. A negative entry, in the amount
repaid, takes care of this.

```{r echo=FALSE}
expenses_doc <- paste(docs, "expenses.csv", sep="")
expenses <- read.csv(file = expenses_doc, header=TRUE, sep=",")
expenses$date <- format(as.Date(expenses$date), date.fmt)
expenses$allowed[is.na(expenses$allowed)] <- 100
expenses$allowed <- expenses$allowed / 100
expenses$expense <- currency(expenses$amount * expenses$allowed, symbol = "€")
expenses$amount = currency(expenses$amount, symbol = "€")
expenses$balance = currency(cumsum(expenses$expense), symbol = "€")
expenses$type = "E"

e <- expenses[, c("date", "description", "amount", "allowed", "expense", "balance")]

formattable(e, list(allowed = percent), align=c("l", "l", "r", "r", "r", "r"))
```
## VAT

VAT is paid to Revenue several times a year, using form VAT3.

The exact amount is the VAT received for Sales minus VAT paid out on Purchases and Director's Expenses.

### Sales (T1)

```{r echo=FALSE}
sales_doc <- paste(docs, "sales.csv", sep="")
sales <- read.csv(file = sales_doc, header=TRUE, sep=",")
sales$date <- format(as.Date(sales$date), date.fmt)
sales$vat.rate <- sales$vat.rate / 100.0
sales$vat.amount <- currency(sales$days * sales$rate * sales$vat.rate, symbol = "€")
sales$balance <- currency(cumsum(sales$vat.amount), symbol = "€")

s <- sales[, c("date", "invoice", "month", "days", "vat.rate", "vat.amount", "balance")]

formattable(s, list(vat.rate = percent), align=c("l", "l", "c", "c", "c", "r", "r"))
```

### Purchases and Expenses (T2)

VAT paid on Purchases from the Bank Account (P) and Director's Expenses (E).

```{r echo=FALSE}
pur_exp <- rbind(purchases, expenses)
pur_exp <- pur_exp[order(as.Date(pur_exp$date, format=date.fmt)),]
#pur_exp <- pur_exp[pur_exp$amount > 0,]
pur_exp$vat.rate[is.na(pur_exp$vat.rate)] <- 0
pur_exp <- pur_exp[pur_exp$vat.rate > 0,]
pur_exp$vat.rate <- pur_exp$vat.rate / 100
pur_exp$vat.amount <- currency(pur_exp$amount / (1 + pur_exp$vat.rate) * pur_exp$vat.rate * pur_exp$allowed, symbol = "€")
pur_exp$amount = currency(pur_exp$amount, symbol = "€")
pur_exp$balance = currency(cumsum(pur_exp$vat.amount), symbol = "€")

p <- pur_exp[, c("date", "description", "type", "amount", "vat.rate", "allowed", "vat.amount", "balance")]
row.names(p) <- NULL

formattable(p, list(vat.rate = percent, allowed = percent), align=c("l", "l", "r", "r", "r", "r", "r"))
```

### Year-to-date
```{r echo=FALSE}
vat3_doc <- paste(docs, "vat3.csv", sep="")
vat3 <- read.csv(file = vat3_doc, header=TRUE, sep=",")
vat3$date <- format(as.Date(vat3$date), date.fmt)
vat3$start <- format(as.Date(vat3$start), date.fmt)
vat3$end <- format(as.Date(vat3$end), date.fmt)
vat3$sales <- currency(vat3$sales, symbol = "€")
vat3$purchases <- currency(vat3$purchases, symbol = "€")
vat3$payable <- currency(vat3$payable, symbol = "€")
vat3$total <- currency(cumsum(vat3$payable), symbol = "€")

formattable(vat3, align=c("l","l","l","r","r","r","r"))
```

### Balance
```{r echo=FALSE}
if (nrow(s) > 0) {
  t1 <- tail(s, 1)$balance - sum(vat3$sales)
} else {
  t1 <- 0
}
if (nrow(p) > 0) {
  t2 <- tail(p, 1)$balance - sum(vat3$purchases)
} else {
  t2 <- 0
}
z <- data.frame(date = c(format(Sys.Date(), date.fmt)), t1, t2, balance = t1-t2)

formattable(z, align=c("l","r","r", "r"))
```
## Trading Details

Trading Details are reported to Revenue at the end of the year, using form RTD.

### Goods and Services
```{r echo=FALSE}
p1 <- currency(sum(sales$amount), symbol = "€")
z1 <- p1
supplies <- data.table(label = c("Std Rate", "Total"), code = c("P1", "Z1"), value = c(p1, z1))
formattable(supplies, align=c("l", "l", "r"))
```
### Other Deductible Goods and Services
```{r echo=FALSE}
e6_rows <- pur_exp[pur_exp$vat.rate == 0,]
e6 <- sum(e6_rows$amount * e6_rows$allowed)
ah6_rows <- pur_exp[pur_exp$vat.rate %in% c(0.135),]
ah6 <- sum(ah6_rows$amount * ah6_rows$allowed)
bh6_rows <- pur_exp[pur_exp$vat.rate %in% c(0.09),]
bh6 <- sum(bh6_rows$amount * bh6_rows$allowed)
r2_rows <- pur_exp[pur_exp$vat.rate %in% c(0.21,0.23),]
r2 <- sum(r2_rows$amount * r2_rows$allowed)
z5 <- sum(e6, bh6, ah6, r2)

deductibles <- data.table(label = c("Exempt", "9%", "13.5%", "Std Rate", "Total"), code = c("E6", "BH6", "AH6", "R2", "Z5"), value = c(e6, bh6, ah6, r2, z5))
formattable(deductibles, align=c("l", "l", "r"))
```
